Converging High-Throughput and High-Performance Computing: A Case Study

### 2017-04-04 overall.tex
Negative spaces fixing
Small iteration 4

Currently, ATLAS deploys 20 instances of the PanDA Broker on 4 Titan's DTNs, 5
instances per DTN. Each broker submits and manages the execution of 15 to 300
ATLAS jobs, one job for each worker node, and a theoretical maximum concurrent
use of 96,000 cores. Since November 2015, PanDA Brokers have operated only in
backfill mode, without a defined time allocation, and running at the lowest
priority on Titan.

@@

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/backfill_consumption.pdf}
    \caption{Titan's total backfill availability: CPU core-hours (gray) and CPU
    core-hours used by ATLAS (blue). GPU core-hours unaccounted for as they
    cannot be used by ATLAS. Efficiency of PanDA Brokers defined as percentage
    of total Titan's backfill availability used by ATLAS (Red labels).}
\label{fig:backfill-utilization}
\end{figure}

PanDA Brokers' efficiency (Fig.~\ref{fig:backfill-utilization}, red labels) is
defined as the fraction of core-hours utilized by the PanDA Brokers to  Titan’s
total backfill availability during the experiment time window. The brokers
reached 18\% average efficiency, with a minimum 7.8\% efficiency on May 2016 and
a maximum 30.9\% efficiency on February 2017. The total backfill availability
was 20.3M in April 2016, and 17.6M in February 2017. This shows that the
measured difference in efficiency did not depend on a comparable difference in
the total backfill availability.

@@

Fig.~\ref{fig:backfill-distrib} shows the distribution of backfill
availability on Titan as a function of number of nodes and the time of their
availability (i.e., walltime). We recorded these data by polling Titan's Moab
scheduler at regular intervals during the experiment window time, while
developing and deploying PanDA Brokers. The mean number of nodes was 691, and
their mean walltime was 126 minutes. Detector simulations of 100 events, enable
to use down to 1 node for 1/2 of the mean walltime of backfill availability. As
such, it offers a good compromise for PanDA Broker efficiency.

\begin{figure*}%[htp]
    \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_wnodes_distribution.pdf}
    \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_walltime_distribution.pdf}
    \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_avail.png}

    \caption{Backfill availability: distribution of number of work nodes (left)
    and of walltime in minutes (center). Scatter plot of the two variables
    (right). 62175 measures: mean number of work nodes available 691; mean
    walltime available 126 minutes.}

@@

Usually, detector simulations performed on the Grid process 10 times the number
of events than those on Titan. This explains the difference between the
percentage of detector simulations and of events computed by those simulations
on Titan. PanDA Broker could fit the number of events to the walltime of the
backfill availability on the base of the distributions of
Fig.~\ref{fig:backfill-distrib}. That specific number of event could then be
pulled from the PanDA Event service~\cite{calafiura2015atlas} and given as input
to one or more simulations. Once packaged into the MPI script submitted to
titan's PBS batch system, these simulations would better fit backfill
availability, increasing the efficiency of PanDA Brokers.

@@

Currently, each PanDA Broker creates, submits, and monitor a single MPI PBS
script at a time. This design is inherited from PanDA Pilot where a single
process is spawn at a time to execute the payload. As a consequence, the
utilization of a larger portion of Titan's total backfill availability depends
on the the number of concurrent PanDA Brokers instantiated on the DTNs: When all
the 20 PanDA Brokers have submitted a MPI PBS Job, further backfill availability
cannot be used.

In September 2016, increasing the number of concurrent PanDA Broker from 4 to 20
markedly improved efficiency (see Fig.~\ref{fig:backfill-utilization}) but
further research is undergoing to understand whether an even larger number
brokers would yield similar results. This research focuses on evaluating the the
overheads of input/output files staging, including its impact on DTNs, and on an
alternative design of PanDA Broker to enable the concurrent submission of
multiple MPI scripts~\cite{barreiro2016panda}.

@@

AthenaMP has an initialization and configuration stage. At initialization time,
AthenaMP is assembled from a large number of shared libraries, depending on the
type of payload that will have to be computed. Once initialized, every algorithm
and service of AthenaMP is configured by a set of Python scripts. Both these
operations result in a large number of read operations on the filesystem shared
between the worker nodes and the DTNs, including the operations required to
access of small python scripts.

Initially, all the shared libraries of AthenaMP and the python scripts for the
configuration stage were stored on the Spider 2 Lustre file system. However, the
I/O patterns of the initialization and configuration stages degraded the
performance of the filesystem. This was addressed by moving the AthenaMP
distribution to a read-only NFS directory, shared among DTNs and worker nodes.
NFS eliminated the problem of metadata contention, improving metadata read
performance from $\approx$6,300 seconds on Lustre to $\approx$1,500 seconds on
NFS.

Once set up, AthenaMP is used to execute 16 concurrent Geant4 simulators on each
of Titan's worker node. Geant4 requires to read events descriptions from a
filesystem and simulate them as they would happen within the ATLAS detector. We
characterized both the compute performance of the simulation and the impact of
acquiring event descriptions on the filesystem.

@@

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx8_tx16_comparison_vsquashed.pdf}

    \caption{Distributions of the time taken to simulate one event when placing
    2 simulations (h1) or 1 simulation (h2) per Titan's CPU. 2 simulation use 16
    cores per node, 1 simulation 8. 50 Events; 1 Titan worker nodes; 16 work
    threads per node; 100 events per node.}

@@

The performance analysis of Titan's AMD CPUs for detector simulations helps also
to compare Titan and Grid site performances. Usually, Grid sites exposes
resources with heterogeneous CPU architectures and a maximum of 8 (virtual)
cores per worker node, while Titan's offer an homogeneous 16 cores architecture.
We used the rate of events processes per minute as a measure of the efficiency
of executing the same detector simulation on both Titan or Grid sites.
Comparisons of the efficiencies of Titan to  the BNL and SIGNET Grid sites,
normalized for 8 cores, show that the effective performance per-core at Titan is
$\approx$0.57 event per minute, roughly 1/2 of BNL and  1/3 of SIGNET
performances.

The differences in performance between Titan and the BNL and SIGNET Grid sites
are due to the FP scheduler competition and the availability of newer
processors. The CPUs at the Grid sites have one FP scheduler per core and are on
average newer than the CPU of Titan. The heterogeneity of the Grid sites' CPUs
explain the higher performance variance compared to the performance consistency
measured on Titan.

>>>>

Currently, ATLAS deploys 20 instances of the PanDA Broker on 4 Titan's DTNs, 5
instances per DTN. Each broker submits and manages the execution of 15 to 300
ATLAS jobs, one job for each Titan's worker node, and a theoretical maximum
concurrent use of 96,000 cores. Since November 2015, PanDA Brokers have operated
only in backfill mode, without a defined time allocation, and running at the
lowest priority on Titan.

@@

\begin{figure}[!t]
    \includegraphics[clip,width=\columnwidth]{figures/backfill_consumption.pdf}
    \vspace{-0.3in}
    \caption{Titan's total backfill availability: CPU core-hours (gray) and CPU
    core-hours used by ATLAS (blue). GPU core-hours unaccounted for as they
    cannot be used by ATLAS. Efficiency of PanDA Brokers defined as percentage
    of total Titan's backfill availability used by ATLAS (Red labels).}
\label{fig:backfill-utilization}
\end{figure}

PanDA Brokers' efficiency (Fig.~\ref{fig:backfill-utilization}, red labels) is
defined as the fraction of core-hours utilized by the PanDA Brokers to  Titan’s
total backfill availability during the experiment time window. The brokers
reached 18\% average efficiency, with a minimum 7.8\% efficiency on May 2016 and
a maximum 30.9\% efficiency on February 2017 (excluding the preliminary results
of March). The total backfill availability was 20.3M in April 2016, and 17.6M in
February 2017. This shows that the measured difference in efficiency did not
depend on a comparable difference in the total backfill availability.

@@

Fig.~\ref{fig:backfill-distrib} shows the distribution of backfill availability
on Titan as a function of number of nodes and the time of their availability
(i.e., walltime). We recorded these data by polling Titan's Moab scheduler at
regular intervals during the experiment window time. The mean number of nodes
was 691, and their mean walltime was 126 minutes. Detector simulations of 100
events, enable to use down to 5/6 of the mean walltime of backfill availability.
As such, it offers a good compromise for PanDA Broker efficiency.

\begin{figure*}[!t]
    \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_wnodes_distribution.pdf}
    \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_walltime_distribution.pdf}
    \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_avail.png}
   \vspace{-0.1in}
    \caption{Backfill availability: distribution of number of work nodes (left)
    and of walltime in minutes (center). Scatter plot of the two variables
    (right). 62175 measures: mean number of work nodes available 691; mean
    walltime available 126 minutes.}

@@

% Usually, detector simulations performed on the Grid process 10 times the
% number of events than those on Titan. This explains the difference between the
% percentage of detector simulations and of events computed by those simulations
% on Titan.
PanDA Broker could fit the number of events to the walltime of each available
backfill slot on the base of the distributions of the time taken by one event to
be simulated. That specific number of event could then be pulled from the PanDA
Event service~\cite{calafiura2015atlas} and given as input to one or more
simulations. Once packaged into the MPI script submitted to titan's PBS batch
system, these simulations would better fit their available backfill slot,
contributing to increase the efficiency of PanDA Brokers.

@@

Currently, each PanDA Broker creates, submits, and monitors a single MPI PBS
script at a time. This design is inherited from PanDA Pilot where a single
process is spawn at a time to execute the payload. As a consequence, the
utilization of a larger portion of Titan's total backfill availability depends
on the the number of concurrent PanDA Brokers instantiated on the DTNs: When all
the 20 PanDA Brokers have submitted a MPI PBS script, further backfill
availability cannot be used.

In September 2016, increasing the number of concurrent PanDA brokers from 4 to
20 markedly improved efficiency (see Fig.~\ref{fig:backfill-utilization}) but
further research is undergoing to understand whether an even larger number of
brokers would yield similar results. This research focuses on evaluating the
overheads of input/output files staging, including its impact on DTNs, and on an
alternative design of PanDA Broker that enables the concurrent submission of
multiple MPI scripts~\cite{barreiro2016panda}.

@@

AthenaMP has an initialization and configuration stage. At initialization time,
AthenaMP is assembled from a large number of shared libraries, depending on the
type of payload that will have to be computed. Once initialized, every algorithm
and service of AthenaMP is configured by a set of Python scripts. Both these
operations result in a large number of read operations on the filesystem shared
between the worker nodes and the DTNs, including the operations required to
access small python scripts.

Initially, all the shared libraries of AthenaMP and the python scripts for its
configuration were stored on the Spider 2 Lustre file system. However, the I/O
patterns of the initialization and configuration stages degraded the performance
of the filesystem. This was addressed by moving the AthenaMP distribution to a
read-only NFS directory, shared among DTNs and worker nodes. NFS eliminated the
problem of metadata contention, improving metadata read performance from
$\approx$6,300 seconds on Lustre to $\approx$1,500 seconds on NFS.

Once initialized and configured, AthenaMP is used to execute 16 concurrent
Geant4 simulators on a Titan's worker node. Geant4 requires to read events
descriptions from a filesystem and simulate them as they would happen within the
ATLAS detector. We characterized both the compute performance of the simulation
and the impact of acquiring event descriptions on the filesystem.

@@

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx8_tx16_comparison_vsquashed.pdf}
   \vspace{-0.3in}
    \caption{Distributions of the time taken to simulate one event when placing
    2 simulations (h1) or 1 simulation (h2) per Titan's CPU. 2 simulation use 16
    cores per node, 1 simulation 8. 50 Events; 1 Titan worker nodes; 16 work
    threads per node; 100 events per node.}

@@

The performance analysis of Titan's AMD CPUs for detector simulations helps also
to compare Titan and Grid site performances. Usually, Grid sites exposes
resources with heterogeneous CPU architectures and a maximum of 8 (virtual)
cores per worker node, while Titan's offer an homogeneous 16 cores architecture.
We used the rate of events processes per minute as a measure of the efficiency
of executing the same detector simulation on Titan or Grid sites. Comparisons of
the efficiencies of Titan to the BNL and SIGNET Grid sites, normalized for 8
cores, show that the effective performance per-core at Titan is $\approx$0.57
event per minute, roughly 1/2 of BNL and  1/3 of SIGNET performances.

The differences in performance between Titan and the BNL and SIGNET Grid sites
are due to the FP scheduler competition and the availability of newer
processors. The CPUs at the Grid sites have one FP scheduler per core and are on
average newer than the CPU of Titan. The heterogeneity of the Grid sites' CPUs
explain the higher performance variance we observed compared to the performance
consistency we measured on Titan.
